% !TeX root = ../main.tex
%------------------------------------------------
% Title page
%------------------------------------------------
\begin{titlepage}
    \centering

    {\Large \textbf{Freie Universität Berlin}}\\[1.5cm]

    {\Large Institute of Meteorology}\\[0.5cm]
    {\Large Department of Earth Sciences}\\[2.0cm]

    {\huge \textbf{Why blockings are the best!}}\\[1.5cm]

    {\Large A cumulative dissertation submitted for the degree of}\\[0.5cm]
    {\Large Doctor of Philosophy (PhD)}\\[1.5cm]

    \begin{tabular}{rl}
        Author: & Henry Leon Schoeller \\
        Supervisor: & Prof.\ Dr.\ Stephan Pfahl \\
        Reviewer: & Prof.\ Dr.\ Gabriele Messori \\
    \end{tabular}

    \vfill

    {\Large June 2026}\\
\end{titlepage}

\cleardoublepage

%------------------------------------------------
% Declaration / Statement of originality
%------------------------------------------------
\chapter*{Declaration}
\addcontentsline{toc}{chapter}{Declaration}

I hereby declare that this dissertation has been composed by myself and that
the work contained herein is my own, except where explicitly stated otherwise.
This dissertation has not been submitted, in whole or in part, for any other
degree or professional qualification.

\vspace{2cm}

\noindent
Berlin, June 2026\\[1.5cm]
\noindent
\textbf{Henry Leon Schoeller}

\cleardoublepage

%------------------------------------------------
% Abstract
%------------------------------------------------
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This dissertation presents a cumulative body of work consisting of three
peer-reviewed research papers. The unifying theme of the dissertation is
briefly described here. The abstract should summarise the research problem,
methodological approach, main contributions of each paper, and the overall
scientific contribution of the thesis.

Keep this concise and factual. Do not include citations unless explicitly
required by your faculty.

\cleardoublepage

%------------------------------------------------
% Acknowledgements (optional but customary)
%------------------------------------------------
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

I would like to thank my supervisor for their guidance and support throughout
the course of this research. I am also grateful to my colleagues, collaborators,
and the members of my research group for their valuable discussions.

Funding sources, institutional support, and personal acknowledgements may be
included here.

\cleardoublepage

%------------------------------------------------
% Table of contents and lists
%------------------------------------------------
\tableofcontents
\cleardoublepage

\listoffigures
\cleardoublepage

\listoftables
\cleardoublepage

%------------------------------------------------
% List of abbreviations (optional)
%------------------------------------------------
\printglossary[type=\acronymtype,title=List of Acronyms]


\cleardoublepage
